Distributional Trees and Forests for Circular Data

### 2019-10-04 pkg/circtree/inst/wind_paper/draft/draft.Rnw
small revisions on reference models and structure of application section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2786 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5bfb7c033c7e056ec72f8786c730574c2280d3dd

\Abstract{
As circular data can be found in many different subject areas accounting for periodicity has 
become an important aspect in establishing appropriate statistical models.
For probabilistic modeling of circular data the von Mises distribution is widely used.
To capture how its parameters change with covariates, a regression tree model
is proposed as an alternative to more commonly-used additive models. The resulting
distributional trees are easy to interpret, can detect non-additive effects, and select
covariates and their interactions automatically. In order to stabelize and regularize
the model an ensemble of distributional trees is combined yielding a distributional
forest which allows for modeling smooth effects as well.
For illustration, hourly wind direction forecasts are obtained at Innsbruck Airport based on a set 
of meteorological measurements from Innsbruck and surrounding weather stations.
}
\Keywords{Distributional Trees; Circular Response; Von Mises Distribution}

@@

In many approaches to model circular data it is assumed that the circular variable of interest 
follows a circular distribution, in particular the von Mises distribution (which will be discussed 
in more detail in Section~\ref{sec:prob_circ}). This was also the basis on which early regression
models for circular data were built on as they regress the distribution parameter(s) on covariates.
One of the first regression models with a circular response variable and linear covariates
was presented by \cite{Gould:1969} where a linear combination of the covariates is employed to model
the circular mean, hence the location parameter of the von Mises distribtuion. 
\cite{Johnson+Wehrly:1978} took this idea one step further 
by plugging in a link function such that the values of the linear predictor are transformed 
to an restricted interval of length $2\pi$. 
This model was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
allowing for the mean direction or the dispersion to be modeled not only in separate models 
but also in a mixed model. However, applying these models has shown to be very challenging, 
mainly due to the complexity encountered in optimizing the log-likelihood function.
Moreover the named approaches and also its further developments 
\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001, Mulder+Klugkist:2017}
are built on additive regression models which may not be able to capture all types of effects.
For example abrupt changes might not be detected properly owing to the fixed structure enforced
by model assumptions such as the type of link function and the type of linear combination.
Therefore, we propose an adaption of regression trees and random forests to circular data by employing 
distributional trees and distributional forests. 
In that way no model assumptions are required as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while combining an ensemble
of trees to a forest model allows for modeling smooth changes as well. 
Therefore, a high level of flexibility is provided while a full distributional model is specified by distributional trees and forests such that a wide range of inference is made available.
These features and the corresponding methodology are presented in Section~\ref{sec:tree_forest} after discussing 
distributional models for circular data in Section~\ref{sec:prob_circ}.
In Section~\ref{sec:wind} the presented models a applied to probabilistic forecasting of wind direction
at the airports Innsbruck and Vienna.

@@

\subsection{Models and Evaluation}
Along with distributional trees for circular data and forests three other models have been applied 
and their performances have been compared based on a circular version of the continuous ranked probability 
score (CRPS).

@@

\paragraph{Climatology}
If there are no climatological changes within a certain number of years it can be expected that 
for a selected day of the year weather conditions will always be the same over all years 
considered.
Again, this idea provides a very simple and stable way of obtaining predictions for a considered 
meteorological variable of interest that can be very useful but its performance can also suffer 
dramatically from a high variance and abrupt changes, especially if they occur frequently during one day.
% TODO: references
Similar to the applied persistence model observations are selected to fit a distributional climatology
model. For a specified day and time point observations from the exact same time point and day but all 
other years contained in the learning data are selected and parameters are estimated following again Section~\ref{sec:prob_circ}.

@@

% TODO: CRPS
\vspace{0.5cm}
Compare fitted models based on a circular version of CRPS ...

>>>>

\Abstract{
As circular data can be found in many different subject areas accounting for periodicity has 
become an important aspect in establishing appropriate statistical models.
For probabilistic modeling of circular data the von Mises distribution is widely used.
To capture how its parameters change with covariates, a regression tree model
is proposed as an alternative to more commonly-used additive models. The resulting
distributional trees are easy to interpret, can detect non-additive effects, and select
covariates and their interactions automatically. In order to stabelize and regularize
the model an ensemble of distributional trees is combined yielding a distributional
forest which allows for modeling smooth effects as well.
For illustration, hourly wind direction forecasts are obtained at two airports in Austria
(Vienna and Innsbruck) based on a set of meteorological measurements from each of the 
airports and surrounding weather stations.
}
\Keywords{Distributional Trees; Circular Response; Von Mises Distribution}

@@

In many approaches to model circular data it is assumed that the circular variable of interest 
follows a circular distribution, in particular the von Mises distribution (which will be discussed 
in more detail in Section~\ref{sec:prob_circ}). This was also the basis on which early regression
models for circular data were built on as they regress the distribution parameter(s) on covariates.
One of the first regression models with a circular response variable and linear covariates
was presented by \cite{Gould:1969} where a linear combination of the covariates is employed to model
the circular mean, hence the location parameter of the von Mises distribtuion. 
\cite{Johnson+Wehrly:1978} took this idea one step further 
by plugging in a link function such that the values of the linear predictor are transformed 
to an restricted interval of length $2\pi$. 
This model was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
allowing for the mean direction or the dispersion to be modeled not only in separate models 
but also in a mixed model. However, applying these models has shown to be very challenging, 
mainly due to the complexity encountered in optimizing the log-likelihood function.
Moreover the named approaches and also its further developments 
\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001, Mulder+Klugkist:2017}
are built on additive regression models which may not be able to capture all types of effects.
For example abrupt changes might not be detected properly owing to the fixed structure enforced
by model assumptions such as the type of link function and the type of linear combination.
Therefore, we propose an adaption of regression trees and random forests to circular data by employing 
distributional trees and distributional forests. 
In that way no model assumptions are required as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while combining an ensemble
of trees to a forest model allows for modeling smooth changes as well. 
In that way, a high level of flexibility is provided while a full distributional model is specified by distributional trees and forests such that a wide range of inference is made available.
These features and the corresponding methodology are presented in Section~\ref{sec:tree_forest} after discussing 
distributional models for circular data in Section~\ref{sec:prob_circ}.
In Section~\ref{sec:wind} the presented models a applied to probabilistic forecasting of wind direction
at the airports of Innsbruck and Vienna.

@@

\subsection{Reference Models}
Along with distributional trees for circular data and forests three other models have been applied 
and their performances have been compared based on a circular version of the continuous ranked probability 
score (CRPS).

@@

\paragraph{Climatology}
If there are no climatological changes within a certain number of years it can be expected that 
for a selected day of the year weather conditions will always be the same over all years 
considered.
Again, this idea provides a very simple and stable way of obtaining predictions for a considered 
meteorological variable of interest that can be very useful but its performance can also suffer 
dramatically from a high variance and abrupt changes, especially if they occur frequently during one day.
% TODO: references
Similar to the applied persistence model observations are selected to fit a distributional climatology
model. For a specified day and time point observations from the exact same time point and the same day 
as well as the 30 surrounding days from all other years contained in the learning data are selected and
parameters are estimated following again Section~\ref{sec:prob_circ}.

@@

\subsection{Evaluation}
Five-fold cross validation (four years as learning data, one year as testing data) ...

Compare fitted models based on a circular version of CRPS ...
